Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales

### 2016-03-18 paper.tex
rho events

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset of \cite{Denton}, which was used in \cite{Takahashi2010}; data are available from 1980 through 1991.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available in the \cite{Denton} dataset from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

In this work, events are defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated in by horizontal dashed lines in their respective panels of Figure~\ref{AllData}.  In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values. \vinote{Not including \req\ events?}

Figure~\ref{ccplot} shows the long-term trends of GOES 6 based $log(\rho_{eq})$ and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows using all non-fill hourly values.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0 and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Our first analysis uses data only from the time interval 1989-1991, which corresponds with the interval used in the analysis in \cite{Takahashi2010}. For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $4\cdot24$ hours were considered before each event and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch day was computed, and these averaging windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.)  Error bars for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values used.

@@

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6, which spans 1983-1991 - a small elevation on the day of the event relative to the day before and day after, with averages after the day of the event being lower than that before.

@@

   Days     p
-1 and 0
-1 and 1
 0 and 1
Results of test on different between means of $\rho_{eq}$ shown in the top panel of \ref{DailyAverage} on days near or on the day of a $D_{st}$ event.  A difference in the means can be rejected with a probability of false rejection of $p$.

   Days     p
-1 and 0
-1 and 1
 0 and 1
Results of test on different between means of $\rho_{eq}$ shown in the bottom panel of \ref{DailyAverage} on days near or on the day of a $D_{st}$ event.

It was found that $\rho_{eq}$ enhancements do not clearly correspond to a signature in $D_{st}$, but that there is a trend for such enhancements to occur when $B_{z}$ in the IMF is positive.  This is consistent with \cite{Takahashi2010} who note that very large increases in $\rho_{eq}$ often corresponded to movement of the plasmasphere to beyond geosynchronous distances.

This method finds 669 periods of $D_{st}$ crossing beneath the selected threshold between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours \vinote{Make sure I defined "duration"}. Figure \ref{HourlyAveragedDstEvents} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{HourlyAveragedDstEvents}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that were used to compute the median. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack.

Figure \ref{HourlyAveragedDstEvents} shows a definite spike in the $Z$ component of the magnetic field, as well as the defined drop in $D_{st}$, but no obvious change in mass density at an hourly timescale. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation since there's no obvious connection on a short timescale. One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long term correlation of density and $D_{st}$.

Two types of events are considered. The first is a drop in $D_{st}$ below the threshold of $-40$~nT.  The second is an increase in $\rho_{eq}$ above $40$~amu/cm$^3$.

Our first analysis uses data only from the time interval 1989-1991.  In this interval, there were 329 $D_{st}$ events and 70 $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event $8\cdot24$ hours were kept after each event and $4\cdot24$ hours before each event.  To compute the epoch averages on a daily time scale, the median value of all measurements for all events centered window of $\pm 12$ hours was computed, and these windows were shifted in increments of $24$~hours. (Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians accross events.)  Error bars for each parameter were computed using the standard deviation of values used in computing the median dividied by the square root of the number of values used.

In the upper panel of \ref{DailyAveragedDstEvents}, $D_{st}$ events are shown.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events corrspond to elevated $\rho_{eq}$, although the magnitude of increase observed here is on the order of $\sim2$~amu/cm$^3$ instead of $\sim10$~amu/cm$^3$ found in \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute its median and error bars.  Note that for $D_{st}$, the number of measurements used in computing the medians is much larger as we use all available $D_{st}$ measurements instead of restricting to only values where a non-fill $\rho_{eq}$ value existed.

Both of the results in \ref{DailyAveragedDstEvents} indicate that there is not a statictically significant change in $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.

Because the dynamics of the magnetopshere take place on a time scale of less than one day, we have also considered hourly averages as shown in the top panel of \ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as \ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of \ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slighly elevated at the time of the threhshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small and in general not statistically significant.

@@

The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack. A subset of longer duration events will be looked at later.

This figure shows a definite spike in the $Z$ component of the magnetic field, as well as the defined drop in $D_{st}$, but no obvious change in mass density at an hourly timescale. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation since there's no obvious connection on a short timescale. One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long term correlation of density and $D_{st}$.

\section{More specific events}
It's hypothesized that progressively picking more specific event criteria will allow for the possibility of more significant results, at the expense of more bias in the selection process and potentially less overall usefulness of the results. That said, the predictability of extreme events is of definite interest, so an attempt has been made to find some reproducible method of prediction. Looking at events that last longer than 12 hours and events with an onset threshhold less than $-80$nT finds both result in a spike in mass density, but also a definite lack of data availability to the point where that spike may be coming from less than five of the total 211 events.

@@

Instead of looking for events based on a threshold of $D_{st}$, we instead looked for a certain amount of change in $\rho_{eq}$ as the basis for event onset. Both raw change and percent change were done in case of bias towards high or low $\rho_{eq}$ periods, respectively, but neither showed any correlation to a change in $D_{st}$.

An alterntive type of event can be defined that corresponds to large increases in $\rho\ $.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\req\ $ above $20$~amu/cm$^2$ defines the start of the event.  Here we see the expected result that such events tend to be associated with positive $B_z^{IMF}$.  To determine if the sign of $B_z^{IMF}$ influences the $\rho\ $ epoch time series, we separated the events by the average of $B_z^{IMF}$ in a time window of $4$~hours centered on the event start time.  The epoch curves for $B_z^{IMF}\geq=0$ (257) events and $B_z^{IMF}\ge=0$ (139) have no statistically signifiant difference.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{paperfigures/F107MD27d-GOES6.eps}
\includegraphics[scale=0.40]{paperfigures/ccplot-GOES6.eps}
\caption{Top: 27-day boxcar averages of $F_{10.7}$ and $\rho_{eq}$ from GOES 6. Bottom: Correlation between $\log(\rho_{eq})$ and $F_{10.7}$ using hour, day, and 27-day averaging intervals; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage

>>>>

The parameters $\rho_{eq}$ and $F_{10.7}$ used in this work are from the dataset of \cite{Denton}, which was used in \cite{Takahashi2010}; data are available from 1980 through 1991 from GOES 2,3,5, 6 and 7.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly average over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

In this work, events are defined to occur when $D_{st}$ or $\rho_{eq}$ crosses a threshold value, as indicated by horizontal dashed lines in their respective panels of Figure~\ref{AllData}.  \vinote{need to put line in rho_eq plot.}

Figure~\ref{ccplot} shows the long-term trends of $log(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0 and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Our first analysis uses data only from the time interval 1989-1991, which corresponds with the interval used in the analysis in \cite{Takahashi2010}. In this interval, there were 329 $D_{st}$ events and 70 $\rho_{eq}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before each event and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch day was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians in 1-day bins and then compute the medians across events.  Error bars for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values used.

@@

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6, which spans 1983-1991 - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.

@@

   Days     p
-1 and 0
-1 and 1
 0 and 1

Caption: Results of test on different between means of $\rho_{eq}$ shown in the top panel of \ref{DailyAverage} on days near or on the day of a $D_{st}$ event.  A difference in the means can be rejected with a probability of false rejection of $p$.

   Days     p
-1 and 0
-1 and 1
 0 and 1

Caption: Results of test on different between means of $\rho_{eq}$ shown in the bottom panel of \ref{DailyAverage} on days near or on the day of a $D_{st}$ event.

Both of the results in \ref{DailyAveragedDstEvents} indicate that there is not a statictically significant change in medians of $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.

Because the dynamics of the magnetopshere take place on a time scale of less than one day, we have also considered hourly averages as shown in the top panel of \ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as \ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of \ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slighly elevated at the time of the threhshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small and in general not statistically significant.  At this timescale no obvious change in mass density at hourly timescale is visible. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation because there is no obvious connection on a short timescale.  One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long-term correlation of density and $D_{st}$.

@@

@@

An alterntive type of event can be defined that corresponds to large increases in $\rho\ $.  In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values. \vinote{Not including \req\ events?}

Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\req\ $ above $20$~amu/cm$^2$ defines the start of the event.  Here we see the expected result that such events tend to be associated with positive $B_z^{IMF}$.  To determine if the sign of $B_z^{IMF}$ influences the $\rho\ $ epoch time series, we separated the events by the average of $B_z^{IMF}$ in a time window of $4$~hours.

Figure~\label{fig:RhoBinned}(a) shows the 
The epoch curves for $B_z^{IMF}\geq=0$ (257) events and $B_z^{IMF}\ge=0$ (139) have no statistically signifiant difference.

@@

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.40]{paperfigures/F107MD27d-GOES6.eps}
\includegraphics[scale=0.40]{paperfigures/ccplot-GOES6.eps}
\caption{Top: 27-day medians of $F_{10.7}$ and $\rho_{eq}$ from GOES 6. Bottom: Correlation between $\log(\rho_{eq})$ and $F_{10.7}$ using medians in hour, day, and 27-day non-overlapping; compare to \cite{Takahashi2010} Fig.~14.}
\label{ccplot}
\end{figure}
\clearpage
